Gravitational-wave sensitivity curves

### 2014-07-03 Introduction.tex
Section 2

\section{Signal analysis}\label{sec:conventions}

Gravitational radiation has two independent polarization states denoted $+$ and $\times$; a general signal can be described as a linear combination of the two polarization states, $h \nobreak=\nobreak A_{+} h_{+}\nobreak +\nobreak A_{\times} h_{\times}$. The sensitivity of a detector to each of these depends upon the relative orientations of the source and detector. The output of a gravitational wave detector, $s(t)$, will contain a superposition of noise, $n(t)$, and possibly a signal, $h(t)$,
\begin{equation} s(t)=n(t)+h(t) \; . \end{equation}
It will be necessary to work with the Fourier transform of the signal, the most common convention used in the GW community is
\begin{eqnarray} x(t)&={\cal{F}}^{-1}\left\{\tilde{x}(f)\right\}&=\int_{-\infty}^{\infty}\mathrm{d}f\;\tilde{x}(f)\exp (2\pi \rmi f t )  \nonumber \\
\tilde{x}(f) &=\,\;{\cal{F}}\left\{x(t)\right\}&=\int_{-\infty}^{\infty}\mathrm{d}t\;x(t)\exp (-2\pi \rmi f t )\; .\end{eqnarray}
For simplicity it is assumed that the noise in the GW detector is stationary and Gaussian (with zero mean), under these assumptions the noise is fully characterised via the \emph{one-sided} noise power spectral density, $S_{n}(f)$,\footnote{An alternative convention is to use the two-sided power spectral density $S^{(2)}_{n}(f)=S_{n}(f)/2$.}, 
\begin{equation}\label{eq:psd} \left<\tilde{n}(f)\tilde{n}^{*}(f')\right>=\frac{1}{2}\delta (f-f')S_{n}(f) \; .\end{equation}
Angled brackets $\left<\ldots\right>$ denote an ensemble average over many realisations of the noise. In reality we have only a single system to work with, so the ensemble average is replaced by a time average. The procedure is to measure the noise over a sufficiently long duration, $T$, and then compute the Fourier transform, $\tilde{n}(f)$, with a frequency resolution $\Delta f = T^{-1}$. This is then repeated many times to give the ensemble average; this procedure assumes the system is ergodic. Since the GW signal and detector output are real it follows that $\tilde{h}(-f)=\tilde{h}^{*}(f)$ and $\tilde{n}(-f)=\tilde{n}^{*}(f)$; therefore, $S_{n}(f)=S_{n}(-f)$. The fact $S_{n}(f)$ is an even function means that Fourier integrals over all frequencies can instead be written as integrals over positive frequencies only (see, for example, equations (\ref{eq:meansquare1}) and (\ref{eq:snrinnerprod})), it is for this reason that $S_{n}(f)$ is called the \emph{one-sided} power spectral density. It is straightforward to show that the power spectral density has the property that integrated over all positive frequencies it gives the mean square noise amplitude in the detector.
\begin{eqnarray}\label{eq:meansquare}
\overline{n(t)^{2}}&=\lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\;\left| n(t) \right|^{2} \nonumber \\
&=\lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi (f-f')t \right)
\end{eqnarray}
Using the property of Fourier transforms, ${\cal{F}}\left\{ n(t+\tau) \right\}(f)=\tilde{n}(f)\exp (-2\pi \rmi f\tau )$, the exponential factors in equation (\ref{eq:meansquare}) may be absorbed as a time translation of the noise by an amount $t$. The noise is stationary and so its long term time average is invariant under this time translation and is given by Equation (\ref{eq:psd}).
\begin{eqnarray}\label{eq:meansquare1}
\overline{n(t)^{2}}& = \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\cal{F}}\left\{ n(\tau-t) \right\}(f){\cal{F}}\left\{ n(\tau-t) \right\}(f')^{*} \nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\nonumber \\
& = \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f)
\end{eqnarray}

Using the property of Fourier transforms, ${\cal{F}}\left\{ n(t+\tau) \right\}(f)=\tilde{n}(f)\exp (-2\pi \rmi f\tau )$, the time average can be turned into an ensemble average and hence the definition of power spectral density in (\ref{eq:psd}) may be used.
\begin{eqnarray}\label{eq:meansquare1}
%\left<n(t)^{2}\right>&=\lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\int_{-T}^{T}\mathrm{d}t\; {\cal{F}}\left\{n(\tau+t)\right\}\left[{\cal{F}}\left\{n(\tau+t)\right\}\right]^{*}\nonumber \\
\overline{n(t)^{2}}& = \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\int_{-T}^{T}\mathrm{d}t\; \tilde{n}(f)\tilde{n}^{*}(f') \nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\nonumber \\
%&= \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta (f-f')\nonumber\\
& = \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f)
\end{eqnarray}
Given a detector output the challenge is to extract the signal. There is a well known solution to this problem which involves constructing a Wiener optimal filter. Let $K(t)$ be the filter function with Fourier transform $\tilde{K}(f)$.
Convolving this with the detector output gives a contribution from the signal and a contribution from the noise
\begin{equation}\label{eq:conv} \left(s*K\right)(\tau)=\int_{-\infty}^{\infty}\mathrm{d}t\;(n(t)+h(t))K(t-\tau)=S+N\; .\end{equation}
The signal contribution, $S$, is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset $\tau$. Since the expectation of pure noise is zero it follows that
\begin{equation} S = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .\end{equation}
The squared contribution from noise, $N^{2}$, is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
N^{2}&=\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left(2\pi\rmi(ft-f't')\right)\nonumber \\
%&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;\frac{1}{2}S_{n}(f)K(t)K^{*}(t')\exp\left(2\pi\rmi f (t-t')\right)\nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
where the definition of $S_{n}(f)$ from (\ref{eq:psd}) has been used. Hence the optimum signal-to-noise ratio (SNR), $\varrho$, is given by
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2}= \frac{(S)^{2}}{N^{2}}= \frac{\left( \frac{1}{2}S_{n}(f)\tilde{K}(f) \Big| \tilde{h} \right)^{2}}{\left( \frac{1}{2}S_{n}(f)\tilde{K}(f) \Big| \frac{1}{2}S_{n}(f)\tilde{K}(f) \right)} \;,
\end{equation}
where the inner product between a pair of signals $A$ and $B$ has been defined
\begin{equation}\label{eq:snrinnerprod} \left( A | B \right) = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter is that which maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy-Schwarz inequality it follows that the optimum filter is
\begin{equation} \tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; . \end{equation}
This is the Wiener filter, it may be multiplied by an arbitrary constant since this will not change the SNR. Notice that in order to construct the Wiener filter it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$. It is common to define the noise amplitude and characteristic strain as
\begin{eqnarray}\label{eq:strains} h_{n}(f)^{2}&=fS_{n}(f) \\
\label{eq:strains1} \label{eq:strainsagain}               h_{c}(f)^{2}&=4f^{2}\left| \tilde{h}(f) \right|^{2} \; , \end{eqnarray}
so that the SNR in (\ref{eq:SNRinnerprod}) may be written compactly as
\begin{equation}\label{eq:hc} 
\varrho^{2}=\int_{0}^{\infty} \mathrm{d}\left(\log f\right)\; \left(\frac{h_{c}(f)}{h_{n}(f)}\right)^{2} \;.
\end{equation}
The strain amplitudes $h_{n}(f)$ and $h_{c}(f)$ are dimensionless, while $S_{n}(f)$ has units of inverse frequency.

Whilst the magnitude of the Fourier transform of the signal, $h(f)=\left|\tilde{h}(f)\right|$, provides a simple quantification of the GW amplitude as a function of frequency it has one main deficiency. For an inspiralling source the instantaneous amplitude can be orders of magnitude below the noise level in the detector; however, as the signal continues over many orbits the SNR can be integrated up to a detectable level. It is desirable to have a quantification of the GW amplitude that accounts for this effect; three such methods are described in the following sections.

\subsection{Characteristic strain}
One sensible choice of quantities to plot on a sensitivity curve is $h_{n}$ for the detector and $h_{c}$ for the source (see figure \ref{fig:hc}). Using this convention, when plotting on a log-log scale, the area between the source and detector curves is related to the SNR via (\ref{eq:hc}). This convention allows the reader to ``integrate by eye" for a given detector to assess the detectability of a given source. Also, when using this convention the reader does not need to worry about the observation time for each source because this is already accounted for in the definition of characteristic strain. An additional advantage of this convention is that the values on the strain axis for the detector curve have a simple physical interpretation, they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis do not directly relate to the amplitude of the waves from the source. Another disadvantage is that the characteristic strain for a monochromatic source is infinite!

\subsection{Power spectral density}\label{sec:psd}
Another common quantity to plot on sensitivity curves is the square root of the power spectral density (see figure \ref{fig:S}), which from (\ref{eq:strains}) is given by
\begin{eqnarray}\label{eq:temp1} \sqrt{S_{n}(f)}&=h_{n}(f)f^{-1/2} \quad \mathrm{for detectors, and by analogy} \\
\sqrt{S_{h}(f)}&=h_{c}(f)f^{-1/2} \quad \mathrm{for sources.}\label{eq:ShforSources}\end{eqnarray}
The power spectral density is the most commonly plotted quantity in sensitivity curves in the literature and it has the nice property, proven in equation (\ref{eq:meansquare1}), that integrated over all positive frequencies it gives the mean square amplitude of the signal in the detector.

\subsection{Energy density}
A third way of describing the detectability of a GW is through the energy carried by the waves, this quantity has the obvious advantage of having a clear physical significance. This quantity is most commonly used in sensitivity curves showing stochastic backgrounds of GWs, see section \ref{sec:stoch}. The energy in GWs is described by the Isaacson stress tensor \citep{MTW}
\begin{equation} T_{\mu\nu}=\frac{c^{2}}{32\pi G}\left<\partial_{\mu}\bar{h}_{\alpha\beta}\partial_{\nu}\bar{h}^{\alpha\beta}\right> \;.\end{equation}
The energy density, $\rho c^{2}$, is given by the $T_{00}$ component of this tensor. Using this together with (\ref{eq:psd}) to calculate the energy density in the GW signal gives
\begin{eqnarray}\label{eq:specNRGdensity}
\rho c^{2}&=\frac{c^{2}}{16\pi G}\int_{-\infty}^{\infty}\mathrm{d}f\;\left(2\pi f\right)^{2}\tilde{h}(f)\tilde{h}^{*}(f)=\int_{0}^{\infty}\mathrm{d}f\;\frac{\pi c^{2}}{4G}f^{2}S_{h}(f)\; ,
\end{eqnarray} 
where the definitions (\ref{eq:ShforSources}) and (\ref{eq:strainsagain}) have been used. Equation (\ref{eq:specNRGdensity}) motivates the definition of the spectral energy density in GWs, $S_{\mathrm{E}}$, the energy per unit volume of space, per unit frequency, \cite{HellingsDowns}
\begin{equation}\label{eq:spectralenergydensity} S_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G} f^{2}S_{h}(f) \; ,\end{equation}
and a corresponding expression for the noise with $S_{n}(f)$. It is usual to define the dimensionless quantity $\Omega_{\mathrm{GW}}$ as the energy density per logarithmic frequency interval normalised to the critical density of the universe
\begin{equation}\label{eq:omega} \Omega_{\mathrm{GW}}(f)=\frac{fS_{\mathrm{E}}(f)}{\rho_{\mathrm{c}}c^{2}}\; . \end{equation}
The critical density of the universe is $\rho_{\mathrm{c}}=3H_{0}^{2}/8\pi G$, where $H_{0}=h_{100}\times 100\, \mathrm{km}\,\mathrm{s}^{-1}\,\mathrm{Mpc}^{-1}$ is the Hubble constant (this equation serves as a definition of $h_{100}$, which here has nothing to do with strain). The most common quantity related to energy density to be plotted on sensitivity curves is $\Omega_{\mathrm{GW}}h_{100}^{2}$ (figure \ref{fig:omega}). This quantity has one aesthetic advantage over the others: it accounts for the fact that there is less energy in low frequency waves of the same amplitude and does not place the sensitivity curves of very low frequency detectors much higher than most ground based detectors.

\subsection{Relating the different descriptions}
The dimensionless energy density in GWs, $\Omega_{\mathrm{GW}}$, the spectral energy density in GWs, $S_{\mathrm{E}}$, the one-sided power spectral density, $S_{h}$, and the characteristic strain are all related via
\begin{equation}\label{eq:omega} 
\Omega_{\mathrm{GW}}(f)=\frac{fS_{\mathrm{E}}(f)}{\rho_{\mathrm{c}}c^{2}}=\frac{\pi}{4G\rho_{\mathrm{c}}}f^{2}h_{c}(f)^{2}=\frac{\pi}{4G\rho_{\mathrm{c}}}f^{3}S_{h}(f)  \; ,
\end{equation}
and corresponding expressions for the noise with $S_{n}(f)$ and $h_{n}(f)$.

>>>>

\section{Signal parametrization}\label{sec:conventions}

\subsection{Signal analysis preliminaries}

Gravitational radiation has two independent polarization states denoted $+$ and $\times$; a general signal can be described as a linear combination of the two polarization states, $h \nobreak=\nobreak A_{+} h_{+}\nobreak +\nobreak A_{\times} h_{\times}$. The sensitivity of a detector to these depends upon the relative orientations of the source and detector. The output of a gravitational wave detector $s(t)$ contains a superposition of noise $n(t)$ and (possibly) the signal $h(t)$,
\begin{equation}
s(t) = n(t)+h(t) \; .
\end{equation}
We will work with the Fourier transform of the signal, using the conventions that
\begin{eqnarray} 
\tilde{x}(f) &= {\mathcal{F}}\left\{x(t)\right\} &= \int_{-\infty}^{\infty}\mathrm{d}t\;x(t)\exp (-2\pi \rmi f t ) \; , \\
x(t) &= {\mathcal{F}}^{-1}\left\{\tilde{x}(f)\right\} &= \int_{-\infty}^{\infty}\mathrm{d}f\;\tilde{x}(f)\exp (2\pi \rmi f t ) \; .
\end{eqnarray}

For simplicity, it is assumed that the noise in the GW detector is stationary and Gaussian (with zero mean); under these assumptions the noise is fully characterised via the one-sided noise power spectral density $S_{n}(f)$, 
\begin{equation}\label{eq:psd}
\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>=\frac{1}{2}\delta (f-f')S_{n}(f) \; ,
\end{equation}
where angle brackets $\left<\ldots\right>$ denote an ensemble average over many noise realisations \citep{Cutler1994}. In reality, we have only a realisation to work with, however the ensemble average can replaced by a time average for stationary stochastic noise. The procedure is to measure the noise over a sufficiently long duration $T$ and then compute the Fourier transform $\tilde{n}(f)$ with a frequency resolution $\Delta f = T^{-1}$; this is repeated many times to give an average. The noise power spectral density $S_{n}(f)$ has units of inverse frequency.

Since the GW signal and detector output are both real it follows that $\tilde{h}(-f)=\tilde{h}^{*}(f)$ and $\tilde{n}(-f)=\tilde{n}^{*}(f)$; therefore, $S_{n}(f)=S_{n}(-f)$. The fact that $S_{n}(f)$ is an even function means that Fourier integrals over all frequencies can instead be written as integrals over positive frequencies only, e.g., (\ref{eq:meansquare1}) and (\ref{eq:snrinnerprod})); it is for this reason that $S_{n}(f)$ is called the \emph{one-sided} power spectral density.\footnote{An alternative convention is to use the two-sided power spectral density $S^{(2)}_{n}(f) = S_{n}(f)/2$.}

When integrated over all positive frequencies the power spectral density gives the mean square noise amplitude. Starting with taking the average of the square of the detector noise:
\begin{eqnarray}\label{eq:meansquare}
\overline{n(t)^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\;\left| n(t) \right|^{2} \nonumber \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft \right)\exp\left(-2\pi \rmi f't \right)
\end{eqnarray}
Using the Fourier property ${\mathcal{F}}\left\{ n(t+\tau) \right\} = \tilde{n}(f)\exp (-2\pi \rmi f\tau )$, the exponential factors may be absorbed as a time-translation of the noise by an amount $t$. The noise is stationary and is invariant under this time translation. We are thus taking the time average...
\cplb{I give up here. This derivation makes no sense in it's current state.}
\begin{eqnarray}\label{eq:meansquare1}
\overline{n(t)^{2}} & = \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\cal{F}}\left\{ n(\tau-t) \right\}(f){\cal{F}}\left\{ n(\tau-t) \right\}(f')^{*} \nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\nonumber \\
& = \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f)
\end{eqnarray}
Using the property of Fourier transforms, ${\cal{F}}\left\{ n(t+\tau) \right\}(f)=\tilde{n}(f)\exp (-2\pi \rmi f\tau )$, the time average can be turned into an ensemble average and hence the definition of power spectral density in (\ref{eq:psd}) may be used.
\begin{eqnarray}\label{eq:meansquare1}
%\left<n(t)^{2}\right>&=\lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\int_{-T}^{T}\mathrm{d}t\; {\cal{F}}\left\{n(\tau+t)\right\}\left[{\cal{F}}\left\{n(\tau+t)\right\}\right]^{*}\nonumber \\
\overline{n(t)^{2}}& = \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\int_{-T}^{T}\mathrm{d}t\; \tilde{n}(f)\tilde{n}^{*}(f') \nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\nonumber \\
%&= \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta (f-f')\nonumber\\
& = \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f)
\end{eqnarray}
\cplb{Try someting based on Wiener--Khinchin theorem?}

Given a detector output, the challenge is to extract the signal. There is a well known solution to this problem which involves constructing a Wiener optimal filter \cplb{[citation needed]}. Let $K(t)$ be the filter function with Fourier transform $\tilde{K}(f)$. Convolving this with the detector output gives a contribution from the signal and a contribution from the noise
\begin{equation}\label{eq:conv}
\left(s*K\right)(\tau) = \int_{-\infty}^{\infty}\mathrm{d}t\;(n(t)+h(t))K(t-\tau) = \mathcal{S} + \mathcal{N} \; .
\end{equation}
The signal contribution $\mathcal{S}$ is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset $\tau$. \cplb{I think this needs expanding.} Since the expectation of pure noise is zero it, follows that
\begin{equation}
\mathcal{S} = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .
\end{equation}
The squared contribution from noise $\mathcal{N}^{2}$ is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
\mathcal{N}^{2} &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left[2\pi\rmi(ft-f't')\right]\nonumber \\
%&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;\frac{1}{2}S_{n}(f)K(t)K^{*}(t')\exp\left(2\pi\rmi f (t-t')\right)\nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
using the definition of $S_{n}(f)$ from (\ref{eq:psd}). Hence the optimum signal-to-noise ratio (SNR) $\varrho$ is given by \cplb{Why is this optimal?}
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2} = \frac{\mathcal{S}^{2}}{\mathcal{N}^{2}}= \frac{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\tilde{h}}^{2}}{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\frac{1}{2}S_{n}(f)\tilde{K}(f)}} \;,
\end{equation}
\cplb{Inner product should be between signals: you don't specify the Fourier transform} where we have introduced the inner product between signal $A$ and $B$ as \citep{Finn1992}
\begin{equation}\label{eq:snrinnerprod} \left( A | B \right) = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy--Schwarz inequality, it follows that the optimum filter is
\begin{equation}
\tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; .
\end{equation}
This is the Wiener filter, it may be multiplied by an arbitrary constant since this does not change the SNR. Using this form for $\tilde{K}(f)$, the squared SNR is
\begin{equation}
\varrho^2 = \int_0^\infty\mathrm{d}f \frac{4| \tilde{h}(f)|^{2}}{S_n(f)} = \innerprod{h}{h}.
\label{eq:traditionalSNR} 
\end{equation}
In order to construct the Wiener filter, it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$.

Whilst the magnitude of the Fourier transform of the signal $|\tilde{h}(f)|$ provides a simple quantification of the GW amplitude as a function of frequency, it has one main deficiency. For an inspiralling source the instantaneous amplitude can be orders of magnitude below the noise level in the detector; however, as the signal continues over many orbits, the SNR can be integrated up to a detectable level. It is useful to have a quantification of the GW amplitude that accounts for this effect; three such methods are described in the following subsections.

\subsection{Characteristic strain}

The characteristic strain $h_c$ is designed to include the effect of integrating an inspiralling signal. Its counterpart for describing noise is the noise amplitude $h_n$. These are defined as
\begin{eqnarray}\label{eq:strain-hc} 
h_\mathrm{c}(f)^{2} &= 4f^{2}\left| \tilde{h}(f) \right|^{2} \; ,\\
h_{n}(f)^{2} &= fS_{n}(f),
\label{eq:strain-hn}
\end{eqnarray}
such that the SNR in (\ref{eq:traditionalSNR}) may be written
\begin{equation}\label{eq:hc} 
\varrho^{2} = \int_{0}^{\infty} \mathrm{d}\left(\log f\right)\; \left[\frac{h_\mathrm{c}(f)}{h_{n}(f)}\right]^{2} \;.
\end{equation}
The strain amplitudes $h_\mathrm{c}(f)$ and $h_{c}(f)$ are dimensionless. Using this convention, when plotting on a log--log scale, the area between the source and detector curves is related to the SNR via (\ref{eq:hc}). This convention allows the reader to integrate by eye to assess the detectability of a given source (see figure \ref{fig:hc}).

An additional advantage of this convention is that the values on the strain axis for the detector curve $h_n(F)$ have a simple physical interpretation: they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis $h_\mathrm{c}(f)$ do not directly relate to the amplitude of the waves from the source. Another disadvantage is that the characteristic strain for a monochromatic source is infinite.

\subsection{Power spectral density}\label{sec:psd}

\cplb{Can we abbreviate to PSD?}

A second commonly used quantity for sensitivity curves is the square root of the power spectral density (see figure \ref{fig:S}). When discussing a detector this is
\begin{equation}\label{eq:temp1}
\sqrt{S_{n}(f)} = h_{n}(f)f^{-1/2} \; ,
\end{equation}
where we have used (\ref{eq:strain-hn}); by analogy, we can define an equivalent for for source amplitudes
\begin{equation}
\sqrt{S_{h}(f)} = h_\mathrm{c}(f)f^{-1/2} = 2 f^{1/2} \left| \tilde{h}(f) \right| \; ,
\label{eq:ShforSources}
\end{equation}
where we have used (\ref{eq:strain-hc}). Both $\sqrt{S_{n}(f)}$ and $\sqrt{S_{h}(f)}$ have units of $\mathrm{Hz^{-1/2}}$. The root power spectral density is the most frequently plotted quantity in the literature.

The power spectral density has the nice property, demonstrated in (\ref{eq:meansquare1}), that integrated over all positive frequencies it gives the mean square amplitude of the signal in the detector. However, in one important regard it is less appealing than characteristic strain: the height of the source above the detector curve is no longer is directly related to the SNR.

\subsection{Energy density}

A third way of describing the amplitude of a GW is through the energy carried by the waves. This has the advantage of having a clear physical significance. The energy density is most commonly used in sensitivity curves showing stochastic backgrounds of GWs (see section \ref{sec:stoch}).

The energy in GWs is described by the Isaacson stress tensor \citep[section 35.15]{MTW}
\begin{equation}
T_{\mu\nu}=\frac{c^{2}}{32\pi G}\left<\partial_{\mu}\bar{h}_{\alpha\beta}\partial_{\nu}\bar{h}^{\alpha\beta}\right> \;,
\end{equation}
where the anle brackets denote averaging over several wavelengths or periods. The energy density $\rho c^{2}$ is given by the $T_{00}$ component of this tensor. Consequently
\begin{eqnarray}
\label{eq:specNRGdensity}
\rho c^{2} &=& \frac{c^{2}}{16\pi G}\int_{-\infty}^{\infty}\mathrm{d}f\;\left(2\pi f\right)^{2}\tilde{h}(f)\tilde{h}^{*}(f) \\*
 &=& \int_{0}^{\infty}\mathrm{d}f\;\frac{\pi c^{2}}{4G}f^{2}S_{h}(f)\; ,
\end{eqnarray} 
where the definition (\ref{eq:ShforSources}) have been used. Equation (\ref{eq:specNRGdensity}) gives the definition of the spectral energy density, the energy per unit volume of space, per unit frequency \citep{HellingsDowns}
\begin{equation}\label{eq:spectralenergydensity}
S_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G} f^{2}S_{h}(f) \; ;
\end{equation}
a corresponding expression for the noise can be formulated by replacing $S_h(F)$ with $S_{n}(f)$.

Cosmological studies often work in terms of the dimensionless quantity $\Omega_{\mathrm{GW}}$, the energy density per logarithmic frequency interval normalised to the critical density of the universe:
\begin{equation}\label{eq:omega}
\Omega_{\mathrm{GW}}(f)= \frac{fS_{\mathrm{E}}(f)}{\rho_{\mathrm{c}}c^{2}} = \frac{\pi}{4G\rho_{c}}f^{3}S_{h}(f) = \frac{\pi}{4G\rho_{c}}f^{2}h_{c}(f)^{2}\; .
\end{equation}
The critical density of the universe is
\begin{equation}
\rho_{\mathrm{c}}=3H_{0}^{2}/8\pi G \;,
\label{eq:crit-density}
\end{equation}
 where $H_{0}$ is the Hubble constant which is commonly parametrized as
\begin{equation}
H_0 =h_{100}\times 100~\mathrm{km\,s^{-1}\,Mpc^{-1}}.
\end{equation}
The reduced Hubble parameter $h_{100}$ has nothing to do with strain. The most common quantity related to energy density to be plotted on sensitivity curves is $\Omega_{\mathrm{GW}}h_{100}^{2}$ (figure \ref{fig:omega}) as this removes sensitivity to the measured value of the Hubble constant.

This quantity has one aesthetic advantage over the others: it accounts for the fact that there is less energy in low frequency waves of the same amplitude and does not place the sensitivity curves of very low frequency detectors much higher than most ground based detectors.

\subsection{Relating the different descriptions}
The dimensionless energy density in GWs $\Omega_{\mathrm{GW}}$, spectral energy density $S_{\mathrm{E}}$, one-sided power spectral density $S_{h}$, characteristic strain $h_\mathrm{c}$ and frequency-domain strain $\tilde{h}(f)$ are related via
\begin{equation}\label{eq:omega} 
H_0^2\Omega_{\mathrm{GW}}(f)= \frac{8 \pi G}{3 c^{2}} fS_{\mathrm{E}}(f) = \frac{2\pi^2}{3} f^3 S_h(f) = \frac{2\pi^2}{3} f^2 h_\mathrm{c}^2(f) = \frac{2\pi^2}{3} f^4 \left|\tilde{h}(f)\right|^2\; ,
\end{equation}
using (\ref{eq:strain-hc}), (\ref{eq:ShforSources}), (\ref{eq:spectralenergydensity}), (\ref{eq:omega}) and (\ref{eq:crit-density}).
Corresponding expressions for the noise are obtained by substituting $S_{n}(f)$ for $S_h(f)$, $h_{n}(f)$ for $h_\mathrm{c}(f)$ and $\tilde{n}(f)$ for $\tilde{h}(f)$.
